Optimistic Atomic Multicast in One Communication Delay

### 2011-07-21 textos/optpaxos/draft_dropbox.tex
ruminating

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\amcast}{\mbox{multicast}}
\newcommand{\amdel}{\mbox{deliver}}
\newcommand{\amdelarg}[1]{\mbox{\amdel({#1})}}
\newcommand{\amcastarg}[1]{\mbox{\amcast({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{opt-deliver}}
\newcommand{\cons}{\mbox{deliver}}
\newcommand{\rmcast}{\mbox{r-mcast}}
\newcommand{\rmdel}{\mbox{r-deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcastarg}[2]{\mbox{\rmcast({#1},{#2})}}
\newcommand{\rmdelarg}[1]{\mbox{\rmdel({#1})}}

@@

%\title{Optimistic Atomic Multicast}
\title{Optimistic Atomic Multicast in One Communication Delay}
%\title{A Quasi-Genuine FIFO Total Order Multicast Primitive}

@@

Processes communicate using reliable multicast, defined by the primitives \rmcast{}$(s,m)$ and \rmdel{}$(m)$, where $s$ is a set of groups and $m$ is a message. Reliable multicast guarantees that (i)~for any process $p$ and any message $m$, $p$ \rmdel{}s $m$ at most once, and only if $p \in s$ and $m$ was previously r-mcast \emph{(uniform integrity)}; (ii)~if a correct process $p$ r-mcasts $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(validity)}; (iii)~if $p$ r-delivers $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(uniform agreement)}; and (iv)~if $p$ r-mcasts $m$ and then $m'$, then no $q$ r-delivers $m'$ without first delivering $m$ \emph{(fifo order)}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $Propose(k, val)$, which proposes a value $val$ for the consensus instance of id $k$ within its group; and also $Decide(k, val)$, which is called when the consensus instance of id $k$ finishes. $Decide(k, val)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $Decide(k, val)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a FIFO reliable multicast primitive \rmcastarg{$m$}{$groupSet$}, which \rmdel{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in FIFO order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $G$ keeps some lists of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{propPending}, containing the messages waiting to be proposed by $p$;
  \item \textit{barPending}, with the messages ready to be \cons{}ed, but which may be waiting for barriers from the groups in $sendersTo(G)$;
  \item \textit{decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{delivered}, which contains the messages that have been \cons{}ed already.
\end{itemize}

@@

The total order delivery of messages in a group can be solved by using consensus. Each consensus instance agrees upon some message set as the next ones to be delivered -- messages within the same set are told apart by the timestamp $m.ts$ applied by the process which created them; to solve timestamp collisions, the unique id of the sender process can be used. If processes send messages to each other using FIFO reliable channels, and after receival, messages are proposed via consensus in the order in which they are received, the FIFO delivery order is ensured.

@@

However, we are using the timestamp given at the creation of a message by its sender. Therefore, it might be the case that, after a message $m'$ has been proposed by a process $p$ in a consensus instance, a message $m : m.ts < m'.ts$ arrives at $p$. If $m$ is delivered with its original timestamp, the timestamp order is violated and there is no sense in using these timestamps for barriers. There are two possible solutions for that: either the message is simply discarded, and no violation to the timestamp order takes place, or we can change the timestamp of $m$ to something greater than the timestamp of $m'$. If uniform validity is to be ensured, discarding $m$ is not an option. As we need, then, to change the value of $m.ts$, two things should be noted: first, as we are using reliable FIFO channels, then the process which sent $m$ is different than that which sent $m'$, so inverting their order does not violate FIFO; finally, increasing a message timestamp must be done with caution, so that messages created by different groups at the same time have roughly the same timestamp and the barrier mechanism is efficient -- if a long sequence of messages have their timestamps increased, the last one of them may wait a long time until all the barriers required to deliver it have arrived.

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

    In this case, $p$ only inserts $m$ in the $barPending$ list for being delivered once all necesarry barriers have arrived.

@@

    In this case, when $m$ is decided, $p$ \rmcast{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is \rmdel{}ed by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. As the sending of these messages from a group to another is done via a FIFO reliable channel, once a message $m$ from another group $G$ is \rmdel{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmdel{}ed.

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $propPending$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $barPending$ list for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $propPending$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $propPending$ that have been already decided are discarded from $propPending$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $barPending$ list. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $barPending$ list. If not, $p'$ inserts $m$ into $barPending$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless FIFO channel\footnote{An ordinary TCP connection would be enough to provide such FIFO lossless channel. Here, we use FIFO reliable multicast.}.

@@

  In this case, when $m$ is received for the first time\footnote{Multiple processes may have sent $m$. To ensure integrity and order, only the first delivery is considered.} by $p$, $m$ is inserted into the $barPending$ list of $p$ and the value of $barrier(G',G)$ is set to $m.ts$.
\end{itemize}

The messages in the \textit{barPending} list are always sorted in ascending order of their timestamps. When the first message $m$ in the $barPending$ list of a process $p \in G$ is such that $m.ts < barrier(G, G')$ for all $G' \in sendersTo(G)$, then $m$ is \cons{}ed by $p$ to the application as the next message. We claim that this delivery respects the FIFO total order\footnote{Proof needed.}.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $propPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $propPending \leftarrow propPending \cup \{m\}$ \label{algline:addtoproppending}
  \ELSIF{$m \notin barPending \wedge m \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$propPending \setminus decided \neq \emptyset$}
	\STATE $propPending \leftarrow propPending \setminus decided$
	\STATE propose($k, propPending$) \label{algline:propose}
	\STATE \textbf{wait until} decide($k, msgSet$)
	\WHILE{$msgSet \setminus decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus decided$ with smallest timestamp
		\STATE let $m'$ be the message in $decided$ with greatest timestamp
		\IF{$m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE
	\STATE $k \leftarrow k + 1$
\ENDWHEN

\blankline
%\WHEN{$\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$} \label{algline:inbarpending}
\WHEN{$barPending \setminus delivered \neq \emptyset$} \label{algline:inbarpending}
	\STATE let $m$ be the message in $barPending \setminus delivered$ with smallest timestamp
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$}
%		\STATE $barPending \leftarrow barPending \setminus \{m\}$ \label{algline:rmfrombp}
		\IF {$m \neq null$}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\ENDIF
		\STATE $delivered \leftarrow delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $propPending \leftarrow propPending \cup \{null\}$
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $propPending$ list only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such list and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$propPending$ queue. % -- to ensure that, once it goes to $propPending$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $decided \leftarrow \emptyset$, $delivered \leftarrow \emptyset$, $propPending \leftarrow \emptyset$, $barPending \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $propPending \leftarrow propPending \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $barPending \leftarrow barPending \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $propPending \leftarrow propPending \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

>>>>

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\amcast}{\mbox{multicast}}
\newcommand{\amdel}{\mbox{deliver}}
\newcommand{\amdelarg}[1]{\mbox{\amdel({#1})}}
\newcommand{\amcastarg}[1]{\mbox{\amcast({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{opt-deliver}}
\newcommand{\cons}{\mbox{deliver}}
\newcommand{\rmcast}{\mbox{fr-mcast}}
\newcommand{\rmdel}{\mbox{fr-deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcastarg}[2]{\mbox{\rmcast({#1},{#2})}}
\newcommand{\rmdelarg}[1]{\mbox{\rmdel({#1})}}

\newcommand{\localmsgs}{messages\text{$_g$}}
\newcommand{\decided}{decided\text{$_g$}}
\newcommand{\stamped}{stamped}
\newcommand{\delivered}{delivered}%{a\mbox{-}delivered}2

@@

%\title{Optimistic Atomic Multicast}
\title{Optimistic Atomic Multicast in One Communication Delay}
%\title{A Quasi-Genuine Fifo Total Order Multicast Primitive}

@@

Processes communicate using fifo reliable multicast, defined by the primitives \rmcast{}$(s,m)$ and \rmdel{}$(m)$, where $s$ is a set of groups and $m$ is a message. Reliable multicast guarantees that (i)~for any process $p$ and any message $m$, $p$ \rmdel{}s $m$ at most once, and only if $p \in s$ and $m$ was previously r-mcast \emph{(uniform integrity)}; (ii)~if a correct process $p$ r-mcasts $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(validity)}; (iii)~if $p$ r-delivers $m$, then eventually all correct processes $q \in s$ r-deliver $m$ \emph{(uniform agreement)}; and (iv)~if $p$ r-mcasts $m$ and then $m'$, then no $q$ r-delivers $m'$ without first delivering $m$ \emph{(fifo order)}.

@@

A more formal description of the protocol is given in Algorithm \ref{algorithm:deliveryminimal}. We consider that three primitives are given: $getTime()$, which returns the current value of the local wallclock; $propose_g(k, v)$, which proposes a value $v$ for the consensus instance of id $k$ within its group; and also $decide_g(k, v)$, which is called when the consensus instance of id $k$ finishes. $decide_g(k, v)$ is called for all the processes of the group that initiated it, when they learn that the value $val$ has been agreed upon in instance of id $k$. For the sake of simplicity, we assume that, for consensus instances within the same group, the values are decided in the same order of the instances id's\footnote{This can be easily done by delaying the callback of $decide_g(k, v)$ while there is some unfinished consensus instance of id $k': k' < k$ from the same group.}. Finally, we also use a fifo reliable multicast primitive \rmcastarg{$m$}{$groupSet$}, which \rmdel{}s $m$ to all the processes in all the groups in $groupSet$ in one communication step, in fifo order (e.g. the one described in \cite{ufrmcast1delta}).

Moreover, each process $p$ of group $g$ keeps some sets of messages:
\begin{itemize}
  %\item \textit{optPending}, which contains the messages waiting to be \mbox{\textit{\opt{}ed}} (optimistically delivered);
  \item \textit{\localmsgs}, containing the %local messages, i.e., the 
  messages that have been \amcast{} by some process of group $g$;
  \item \textit{\decided}, which contains the messages that have already been proposed and decided within $group(p)$;
  \item \textit{\stamped}, containing the messages sent to $g$ by any group; a final timestamp has already been assigned to each of these messages, although, in order to be \amdel{}ed, some of them may still need barriers from the groups in $sendersTo(G)$;
  \item \textit{\delivered}, which contains the messages that have been already \amdel{}ed by $p$.
\end{itemize}

@@

The total order delivery of messages in a group can be solved by using consensus. Each consensus instance agrees upon some message set as the next ones to be delivered -- messages within the same set are told apart by the timestamp $m.ts$ applied by the process which created them; to solve timestamp collisions, the unique id of the sender process can be used. If processes send messages to each other using fifo reliable channels, and after receival, messages are proposed via consensus in the order in which they are received, the fifo delivery order is ensured.

@@

However, we are using the timestamp given at the creation of a message by its sender. Therefore, it might be the case that, after a message $m'$ has been proposed by a process $p$ in a consensus instance, a message $m : m.ts < m'.ts$ arrives at $p$. If $m$ is delivered with its original timestamp, the timestamp order is violated and there is no sense in using these timestamps for barriers. There are two possible solutions for that: either the message is simply discarded, and no violation to the timestamp order takes place, or we can change the timestamp of $m$ to something greater than the timestamp of $m'$. If uniform validity is to be ensured, discarding $m$ is not an option. As we need, then, to change the value of $m.ts$, two things should be noted: first, as we are using reliable fifo channels, then the process which sent $m$ is different than that which sent $m'$, so inverting their order does not violate fifo; finally, increasing a message timestamp must be done with caution, so that messages created by different groups at the same time have roughly the same timestamp and the barrier mechanism is efficient -- if a long sequence of messages have their timestamps increased, the last one of them may wait a long time until all the barriers required to deliver it have arrived.

@@

  When a process in $G$ \amcast{}s $m$, it is first \rmcast{} to the other processes of $G$. Then, when $m$ is \rmdel{}ed at $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $\localmsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in $I_{k+1}$.

  Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $\stamped$ set for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t_b$, such that $t_b \geq m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.

@@

    In this case, $p$ only inserts $m$ in the $\stamped$ set for being delivered once all necesarry barriers have arrived.

@@

    In this case, when $m$ is decided, $p$ \rmcast{}s $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is \rmdel{}ed by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $\stamped$ set. If not, $p'$ inserts $m$ into $\stamped$ and adjusts $barrier(G, G')$ to $m.ts$. As the sending of these messages from a group to another is done via a fifo reliable channel, once a message $m$ from another group $G$ is \rmdel{}ed, every message $m' : m'.ts < m.ts$ from $G$ has also already been \rmdel{}ed.

  \end{itemize}
%   
%   In this case, when $m$ is received by $p$, $p$ checks whether the latest consensus instance $I_{k}$ in which it participated, or is trying to start, has already been decided -- if not, $p$ enqueues $m$ in a $\localmsgs$ queue as the next message being proposed by it in the consensus instance $I_{k+1}$, so other tasks can keep being executed. Then, once $I_{k}$ has been decided, $p$ may start a new instance. Before that, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in $I_{k+1}$. Once $m$ is decided, it is not immediately delivered to the application. Instead, $p$ checks whether some message $m_{prv} : m_{prv}.ts \geq m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, $m$ is inserted into a $\stamped$ set for later being delivered, which will happen once every group $G'$ in $sendersTo(G)$ has already sent a message $barrier(G',G) = t$, such that $t > m.ts$. This is done because there could be a message $m'$ yet to come from another group $G'$, such that $m'.ts < m.ts$.
% 
%   \item When $m$ is originated in $G$, but it has at least one group other than $G$ as a destination:
%   
%   In this case, when $m$ is received, $p$ tries to initiate a consensus instance within $G$ to decide $m$. If $p$ cannot start the proposal now, $m$ is enqueued in $\localmsgs$ for being proposed later along with other pending messages. Then, once $p$ may start a new instance, all messages in $\localmsgs$ that have been already decided are discarded from $\localmsgs$. The rest is proposed as a batch in the new consensus instance. Once any message $m$ is decided, $p$ checks whether some message $m_{prv} : m_{prv}.ts > m.ts$ has been decided previously. If that is the case, the value of $m.ts$ is changed to a value greater than the timestamp of any other message previously decided within $G$. Then, if $G \in m.dst$, $m$ is inserted in the $\stamped$ set. Besides, when $m$ is decided, $p$ sends $m$ to every \mbox{$p' \in (m.dst \setminus \{G\})$}. When $m$ is received by each $p' \in G'$, $p'$ checks whether it has ever inserted $m$ in its own $\stamped$ set. If not, $p'$ inserts $m$ into $\stamped$ and adjusts $barrier(G, G')$ to $m.ts$. To ensure that, once a message $m$ is received from another group $G$, every message $m' : m'.ts<m.ts$ also from $G$ has already been received, every message is sent through a lossless fifo channel\footnote{An ordinary TCP connection would be enough to provide such fifo lossless channel. Here, we use fifo reliable multicast.}.

@@

  In this case, when $m$ is received for the first time\footnote{Multiple processes may have sent $m$. To ensure integrity and order, only the first delivery is considered.} by $p$, $m$ is inserted into the $\stamped$ set of $p$ and the value of $barrier(G',G)$ is set to $m.ts$.
\end{itemize}

The messages in the \textit{\stamped} set are always handled in ascending order of their timestamps. Let $m$ be the message with lowest timestamp in the $\stamped$ set of a process $p \in g$. When $m.ts < barrier(g, h)$ for all $h \in sendersTo(g)$, then $m$ is \amdel{}ed by $p$ to the application as the next message. We proved that this delivery respects the fifo total order.

@@

\blankline
\INDENT{Initialization}
  \STATE $k \leftarrow 0$, $\localmsgs \leftarrow \emptyset$, $\decided \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$
  \INDENT{\textbf{for all} $h \in sendersTo(g)$ \textbf{do}}
    \STATE $barrier(h) \leftarrow (0,0)$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\WHEN{\rmdelarg{$m$}}
  \IF{$g = m.src$}
    \STATE $\localmsgs \leftarrow \localmsgs \cup \{m\}$ \label{algline:addtogroupmsgs}
  \ELSIF{$m \notin \stamped$} \label{algline:notbptwice}
    \STATE $\stamped \leftarrow \stamped \cup \{m\}$ \label{algline:insbp1}
    \STATE $barrier(m.src) \leftarrow m.ts$ \label{algline:incbar}
  \ENDIF
\ENDWHEN

\blankline
\WHEN{$\localmsgs \setminus \decided \neq \emptyset$}
  \STATE $k \leftarrow k + 1$
	\STATE $undecided \leftarrow \localmsgs \setminus \decided$ \label{algline:undecided}
	\STATE propose$_g$($k, undecided$) \label{algline:propose}
	\STATE \textbf{wait until} decide$_g$($k, msgSet$) \label{algline:decide}
	\WHILE{$msgSet \setminus \decided \neq \emptyset$}
		\STATE let $m$ be the message in $msgSet \setminus \decided$ with smallest timestamp \label{algline:decidedloopbegin}
		\STATE let $m'$ be the message in $\decided$ with greatest timestamp
		\IF{$m'.ts > m.ts$}
			\STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:changetsafterdecision}
		\ENDIF     
		\IF{$g \in m.dst$} \label{algline:checkcons}
			\STATE $\stamped \leftarrow \stamped \cup \{m\}$  \label{algline:insbp2}
		\ENDIF
		\STATE $\decided \leftarrow \decided \cup \{m\}$  \label{algline:addtodecided}
		\STATE \rmcastarg{$m.dst \setminus \{g\}$}{$m$} \label{algline:rmothers}
	\ENDWHILE

\ENDWHEN

\blankline
%\WHEN{$\exists m \in \stamped : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in \stamped : m'.ts < m.ts$} \label{algline:in\stamped}
\WHEN{$\stamped \setminus \delivered \neq \emptyset$} \label{algline:inammulticast}
	\STATE let $m$ be the message in $\stamped \setminus \delivered$ with smallest timestamp \label{algline:condel1}
	\IF{$\forall h \in sendersTo(g): m.ts < barrier(h)$} \label{algline:condel2}
%		\STATE $\stamped \leftarrow \stamped \setminus \{m\}$ \label{algline:rmfrombp}
		  \STATE \amdelarg{$m$} \label{algline:consdeliver}
		\STATE $\delivered \leftarrow \delivered \cup \{m\}$ \label{algline:addtodelivered}
	\ENDIF
\ENDWHEN
\blankline

@@

\blankline
\INDENT{\textit{When $\exists h: g \in sendersTo(h) \wedge getTime() - lastBarrierCreated(h) > barrierThreshold$}}
  \STATE $null \leftarrow$ empty message
  \STATE $null.ts \leftarrow (lastBarrierCreated(h) + barrierThreshold, 0)$ \label{algline:samenullid}
  \STATE $null.src \leftarrow g$
  \STATE $null.dst \leftarrow \{h\}$ 
  \STATE $\localmsgs \leftarrow \localmsgs \cup \{null\}$ \label{algline:enqnull}
  \COMMENT{saving that nothing was sent until null.ts}
\ENDINDENT

@@

It is necessary to guarantee that a $null$ message will eventually be proposed, decided, and a barrier will be sent to some group which might be needing it, so that progression is guaranteed. Therefore, this kind of messages are created by every process in the group, since if any one of them does not have it, such message might never be decided. %Besides, they are excluded from the $\localmsgs$ set only when decided in the group (l. \ref{algline:nullstays} of Algorithm \ref{algorithm:deliveryminimal}). Since they are obviously never delivered to the application, they can remain in such set and be decided even after some message with higher timestamp is decided. Not removing them unless they are decided ensures that they will be sent and that other groups will be able to increase their barrier values. % On the other hand, this brings another problem: if every process in $G$ creates a different $null$ message, many $null$ messages might have to be decided before a meaningful message.
To prevent the multiple $null$ messages -- created by different processes within the group -- of being decided, they could be created in a way such that the different processes can somehow figure out that two different $null$ messages are equivalent\footnote{This could be done by assuming no timestamp collisions and by using them to uniquely identify messages. Then, only one of the messages created with the same timestamp (l. \ref{algline:samenullid} of Algorithm \ref{algorithm:nullperiodic}) would be decided.}.

@@

When the process $p'$ of some group $G'$ receives a barrier request for a message $m$, %such that $G' \in blockers(m)$,
$p'$ knows that there are other groups depending on the barrier of $G'$ to deliver $m$. For that reason, it will immediately create a $null$ message with a timestamp equal to $m.ts$ and with $null.dst = m.dst$ $\cap$ $receiversFrom(G')$. Then, $p'$ will insert such message in its %$optPending$
$\localmsgs$ queue. % -- to ensure that, once it goes to $\localmsgs$, any message \mbox{$m': m'.ts < null.ts \wedge m'.src = G'$} is already there. As soon as $null.ts > now - w(p')$, the 
Once the $null$ message is proposed and decided in $G'$, each process of $G'$ will send it to every process in each group \mbox{$G \in m.dst \cap receiversFrom(G')$}. This way, any group which was waiting for a barrier from $G'$ to deliver $m$ will be able to do so as soon as it receives such $null$ message. The new delivery algorithm would be as described in \mbox{Algorithm {\ref{algorithm:nullondemand}}}.

@@

\blankline
\INDENT {Initialization}
  \STATE $k \leftarrow 0$, $nextProp \leftarrow 0$, $\decided \leftarrow \emptyset$, $\delivered \leftarrow \emptyset$, $\localmsgs \leftarrow \emptyset$, $\stamped \leftarrow \emptyset$
  \INDENT{\textbf{for all} $G' \in sendersTo(G)$ \textbf{do}}
    \STATE $barrier(G',G) \leftarrow -\infty$ 
  \ENDINDENT
\ENDINDENT

@@

\blankline
\INDENT{\textit{When} \rmdelarg{$m'$}}
  \IF{$G = m'.src$}
    \STATE $\localmsgs \leftarrow \localmsgs \cup \{m'\}$
  \ELSIF{$m' \notin \stamped \wedge m' \notin \delivered$}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
    \IF {$G \in m'.dst$}
      \STATE $\stamped \leftarrow \stamped \cup \{m'\}$
    \ENDIF
    \IF{$G \in blockers(m') \wedge m' \neq null $}
      \STATE $null \leftarrow$ empty message
      \STATE $null.src \leftarrow G$
      \STATE $null.ts \leftarrow m'.ts$ \label{algline:nulltsmts}
      \STATE $null.dst \leftarrow m'.dst \cap receiversFrom(G)$
      \STATE $local\_msgs \leftarrow local\_msgs \cup \{null\}$
    \ENDIF
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in \localmsgs \wedge nextProp = k$}}
%  \STATE $\localmsgs \leftarrow \localmsgs \setminus \{m\}$
%  \IF {$m \notin \decided \wedge \nexists m' \in \decided: m'.ts > m.ts$}
    \STATE $\localmsgs \leftarrow \localmsgs \setminus \decided$%(\decided \cup \{m' : \exists m'' \in \decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:nullstays}
    \IF {$\localmsgs \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE propose$_g$($k$,$\localmsgs$)
    \ENDIF
%  \ENDIF
\ENDINDENT
